### 2025-04-25 code/Draft_script.qmd
initial commit from lab desktop

---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: Md Shakir Moazzem, Umar Munir
---

# Introduction

This script contains the data wrangling steps for the final project.

# Setup

##Loading packages

The following code chunk will load necessary packages.

```{r Setup, message=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("janitor")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("readr")
#install.packages("lubridata")
#install.packages("stringr")

# Loading packages

library(tidyverse)
library(readxl) # to read excel files
library(janitor) # to clean data; helps fix and standardize the column names
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(lubridate)
library(stringr)

```

## Reading data

The following code chunk will read the csv files for the 3 training data sets

```{r training data import, message=F, warning=F}

#reading the csv files for the 3 training data sets

trait_df <- read_csv("../data/training/training_trait.csv") 
meta_df  <- read_csv("../data/training/training_meta.csv")
soil_df  <- read_csv("../data/training/training_soil.csv")

```

```{r}

# Function to transform yield to 15.5% moisture
adjust_yield <- function(yield_mg_ha, grain_moisture) {
  yield_mg_ha * (100 - grain_moisture) / (100 - 15.5)
}

```

```{r clean_and_summarize_trait, message=F, warning=F}
library(stringr)

trait_avg_df <- trait_df %>%
  select(-block) %>%
  # 1) remove all lowercase letters from site
  # 2) keep only text before any dash
  # 3) keep only text before any underscore
  mutate(
    site = str_remove_all(site, "[a-z]"),
    site = str_replace(site, "-.*$", ""),
    site = str_replace(site, "_.*$", "")
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted   = first(date_planted),
    date_harvested = first(date_harvested),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy      = yday(planting_date),
    harvest_doy    = yday(harvest_date),
    season_len     = harvest_doy - plant_doy
  ) %>%
  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested) %>%
  ungroup()

trait_avg_df

```

```{r exporting wrangled data set}

write_csv(trait_avg_df,
          "../data/training_trait_clean.csv")
```

The following code chunk will conduct data wrangling on  "training_soil.csv"

```{r data_wrangling_training_soil, message=F, warning=F}

library(stringr)

soil_clean <- soil_df %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  )
```

```{r exporting wrangled data set}

write_csv(soil_clean,
          "../data/training_soil_clean.csv")
```

The following code chunk will conduct data wrangling on  "training_meta.csv".

```{r clean_meta_sites}

meta_clean <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  select(-previous_crop) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

```

```{r exporting wrangled data set}

write_csv(meta_clean,
          "../data/training_meta_clean.csv")
```

```{r}

merged_clean  <- trait_avg_df %>%
  left_join(soil_clean, by = c("year", "site")) %>%
  left_join(meta_clean, by = c("year", "site"))

summary(merged_clean)
```

# Exporting

The following code chunk will export the wrangled data into a .csv.

```{r exporting wrangled data set}

write_csv(merged_clean,
          "../data/training_merged_clean.csv")
```

# End of Data Wrangling

#Start of Open Source Daymet weather data download

The following code chunk will load necessary packages.

```{r Setup, message=F, warning=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("sf") #to manipulate vector geospatial files
#install.packages("daymetr") #to retrieve data from daymet database website through R
#install.packages("remotes") #to install R packages that are not available on CRAN, and available on GitHub
#remotes::install_github("ropensci/USAboundaries") 
#remotes::install_github("ropensci/USAboundariesData")

# Loading packages

library(tidyverse) #need to load "tidyverse" package at first
library(sf) # for US map #to manipulate vector geo-spatial points
library(daymetr) #to retrieve data from daymet database website through R
library(remotes)
library(USAboundaries) # for US state boundaries
library(USAboundariesData)

```

```{r create map of USA and add points, message=F, warning=F}

states <- us_states() %>% 
  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii) from the rows of "states" object i.e., to keep all states that are NOT "PR", "AK", and "HI" #we use "%in%" to filter (i.e., work on) more than one state/entry #if we wanted to filter just 1 state, we would use == sign e.g., !(state_abbr == c("PR"))

ggplot() +
  geom_sf(data = states) + #"geom_sf()" is used to plot "sf" object, which we just created above as "states" object; plots all states and territories of USA
  geom_point(data = merged_clean,
             aes(x = lon, #"Longitude" goes on longitude
                 y = lat) #"Latitude" goes on latitude
             ) +
  labs(
    title = "Corn Trial Site Locations (2014–2023)",
    x     = "Longitude",
    y     = "Latitude"
  )

```

```{r}

# Define Daymet bounding box (WGS-84)
min_lat <- 14.53
max_lat <- 52.00
min_lon <- -131.104
max_lon <- -52.95

# Filter merged to Daymet’s valid range, dropping any NA coords
merged_daymet <- merged_clean %>%
  filter(
    !is.na(lat),
    !is.na(lon),
    lat  >= min_lat,
    lat  <= max_lat,
    lon  >= min_lon,
    lon  <= max_lon
  )

# Report how many rows remain (and were dropped)
message("Rows kept: ", nrow(merged_daymet), 
        " (dropped: ", nrow(merged_clean) - nrow(merged_daymet), ")")

```

```{r unique_site_years_with_coords}

# Extract unique combinations of year, site, and their coordinates
site_year_df <- merged_daymet %>%
  select(year, site, lon, lat) %>%  # include longitude and latitude
  distinct() %>%                     # one row per unique combination
  arrange(year, site)

# Inspect the result
site_year_df

```

The following code chunk will download the weather data for all unique combinations of year, site, and their coordinates in the "site_year_df" object.

```{r}

weather_daymet_all <- site_year_df %>% 
  mutate(weather = pmap(list(.y = year, 
                             .site = site, 
                             .lat = lat, 
                             .lon = lon), 
                        function(.y, .site, .lat, .lon) 
                          download_daymet( 
                            site = .site, 
                            lat = .lat, #specifying ".lat" placeholder for "lat = " argument
                            lon = .lon, #specifying ".lon" placeholder for "lon = " argument
                            start = .y, 
                            end = .y, 
                            simplify = T,
                            silent = T) %>% #end of " download_daymet()" function
                          rename(.year = year,
                                 .site = site) 
                        ))

```

The following code chunk will unnest the weather column.

```{r}

weather_daymet_unnest <- weather_daymet_all %>%
  select(year, site, weather) %>% 
  unnest(weather) %>% 
  pivot_wider(names_from = measurement, 
              values_from = value) %>% 
  janitor::clean_names() #

weather_daymet_unnest

```

# Merging the weather data retrieved from Daymet to the "merged_clean" data

```{r merge_weather_with_data}

# Join daily weather onto your cleaned trial data by year & site

daymet_all_unnest <- merged_clean %>%
  left_join(
    weather_daymet_unnest,
    by = c("year", "site")
  )

daymet_all_unnest

```

## Exporting the merged data

```{r}

write_csv(daymet_all_unnest,
          "../data/merged_fieldweatherdata.csv"
          )

```

# End of retreiving weather data from Daymet

# Start of feature engineering

## Setup

```{r}
#| message: false
#| warning: false

install.packages("ggridges")

library(ggridges)
library(tidyverse)

```

```{r}

fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")

fieldweather

```

```{r}

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, site, hybrid, lon, lat, yday, 
                yield = adjusted_yield, 
                planting.date = planting_date, 
                harvest.data = harvest_date, 
                planting.doy = plant_doy, 
                harvest.doy = harvest_doy, 
                season.length = season_len, 
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

fe_month
```

Now, let's summarize daily weather variables based on month.

```{r fe_month_sum}

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

fe_month_sum

```

Let's check tmax.c and prcp.mm for the first site-year and month.

[Note: the following code chunk is for double checking to make sure that we did everything okay the way we intended]

```{r}

fe_month %>%
  filter(year == 2014 & 
           site == "DEH1" &
           hybrid == "B37/MO17" &
           month_abb == "Jan") %>%
  summarise(tmax.c = mean(tmax.c),
            prcp.mm = sum(prcp.mm))

```

The code below will put the month as part of the column name.

```{r fe_month_sum_wide}

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(5:ncol(.), ~round(., 1) )) %>%
  select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

fe_month_sum_wide

```

# EDA round 2  
Let's make a ridge plot to visualize the distribution of one variable over months.

```{r, message=FALSE, warning=FALSE}

#install.packages("ggridges")
library(ggridges) #really powerful package to plot distributions e.g., density plot

ggplot(data = fe_month_sum,
       aes(x = mean_tmax.c,
           y = month_abb,
           fill = stat(x) #fill = stat(x): specific to "ggridges"
           )
       ) +
  geom_density_ridges_gradient(scale = 3,
                               rel_min_height = 0.01) + #"the 2nd argument is"rel_min_height = 0.01": to cut down the tails of the distribution to look nice
  scale_fill_viridis_c(option = "C") + #"viridis"scale_fill_viridis_c()": color-blind-friendly color scale #argument (option = "C") to change the color to "magma" because magma is weather related; "option = " varies from A to F
  theme(legend.position = "none") #"theme(legend.position = "none")": to remove the legend from the graph

```

Now let's do that for all weather variables [ we will automate using map2() to conduct iteration ]

```{r}

finalplots <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% #we need to do pivot_longer() to iterate over the weather variables
  group_by(name) %>%
  nest() %>% #for iteration, we use "group_by()" followed by "nest()" in combo
  #we will use map2() since we need to iterate over 2 columns
  mutate(plot = map2(data, name, #map2() takes 2 arguments: the 1st argument becomes .x, the 2nd argument becomes .y #we must use map2() with a mutate() at first for iteration #"map2(data, name" : we want to iterate over "name" for "data"
                     ~ ggplot( data = .x, # .x represent "data" in the map2() function, so we need to use data = .x [= "data" from map2() ] to feed the "data" from map2() function as the data of ggplot() #.x is the iterating column of map2() that is a place holder for the 1st argument of map2()
       aes(x = value,
           y = month_abb,
           fill = stat(x)
           )
       ) +
  geom_density_ridges_gradient(scale = 3,
                               rel_min_height = 0.01) + 
  scale_fill_viridis_c(option = "C") + 
  theme(legend.position = "none") +
  labs(x = .y) #to rename the x-axis as the variable name # .y is the placeholder for "name" (i.e., variable name) in map2() function
                     ))

finalplots

```

```{r}
#| message: false

finalplots$plot #$plot to print all the ggplots for each variables

```

#Export weather_monthsum.csv

```{r}

write_csv(fe_month_sum_wide,
          # "../data/weather_monthsum.csv")

```

### 2025-04-25 code/Draft_script.qmd
d

@@
 ---
+
+# Start of data wrangling
 
@@
 
+## EDA
+
+```{r}
+
+summary(trait_df)
+
+```
+```{r}
+
+summary(meta_df)
+
+```
+
+```{r}
+
+summary(soil_df)
+
+```
+
+# Data wrangling on "trait_df"
+
+The code below creates a function to adjust the yield_mg_ha for 15.5% grain moisture.
+
 ```{r}
@@
 
+The code below conducts data wrangling on "trait_df"
+
 ```{r clean_and_summarize_trait, message=F, warning=F}
-library(stringr)
 
-trait_avg_df <- trait_df %>%
+trait_clean <- trait_df %>%
   select(-block) %>%
-  # 1) remove all lowercase letters from site
-  # 2) keep only text before any dash
-  # 3) keep only text before any underscore
   mutate(
-    site = str_remove_all(site, "[a-z]"),
-    site = str_replace(site, "-.*$", ""),
-    site = str_replace(site, "_.*$", "")
+    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
+    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
+    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
   ) %>%
@@
     harvest_doy    = yday(harvest_date),
-    season_len     = harvest_doy - plant_doy
   ) %>%
-  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested) %>%
+  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested, -planting_date, -harvest_date) %>%
   ungroup()
@@
 
-write_csv(trait_avg_df,
-          "../data/training_trait_clean.csv")
+#write_csv(trait_avg_df,
+          #"../data/training_trait_clean.csv")
 ```
+
+# Data wrangling on "soil_df"
 
@@
     site = str_replace(site, "_.*$", "")
-  )
+  ) %>%
+  ungroup()
+
+summary(soil_clean)
 ```
@@
 
-write_csv(soil_clean,
-          "../data/training_soil_clean.csv")
+#write_csv(soil_clean,
+          #"../data/training_soil_clean.csv")
 ```
@@
   ) %>%
-  select(-previous_crop) %>%
   mutate(
@@
 
+The code below will create categories for previous crops.
+
+```{r}
+library(stringr)
+
+meta_clean <- meta_clean %>%
+  mutate(
+    # 1) lowercase & trim
+    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
+    # 2) recode into your desired levels
+    prev_crop = case_when(
+      prev_crop_raw == "cotton"                                   ~ "cotton",
+      prev_crop_raw == "peanut"                                   ~ "peanut",
+      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
+      prev_crop_raw == "sorghum"                                  ~ "sorghum",
+      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
+      TRUE                                                         ~ "others"
+    ) %>%
+    # 3) make it a factor with the exact ordering you want
+    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
+  ) %>%
+  select(-previous_crop, -prev_crop_raw) %>%
+  clean_names()
+
+meta_clean
+
+```
+
+```{r}
+summary(meta_clean)
+```
+
 ```{r exporting wrangled data set}
 
-write_csv(meta_clean,
-          "../data/training_meta_clean.csv")
+#write_csv(meta_clean,
+          #"../data/training_meta_clean.csv")
 ```
@@
 
-write_csv(merged_clean,
-          "../data/training_merged_clean.csv")
+#write_csv(merged_clean,
+          #"../data/training_merged_clean.csv")
 ```
@@
 
-write_csv(daymet_all_unnest,
-          "../data/merged_fieldweatherdata.csv"
+#write_csv(daymet_all_unnest,
+          #"../data/merged_fieldweatherdata.csv"
           )
@@
 
-fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")
+#fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")
 
-fieldweather
+#fieldweather
 
@@
 
-write_csv(fe_month_sum_wide,
+#write_csv(fe_month_sum_wide,
           # "../data/weather_monthsum.csv")
@@
 ```
+
+# Feature engineering "planting stages" and "harvest stages":
+
+The code chunk below is to feature engineer planting and harvest stages based on the following cut-offs for planting day of the year (plant_doy) and harvest day of the year (harvest_doy):
+
+Planting stages:
+ - early: plant_doy ≤ 120
+ - normal: plant_doy = 121–150
+ - late: plant_doy > 150
+
+Harvest stages:
+ - early: harvest_doy ≤ 275
+ - normal: harvest_doy = 276–305
+ - late: harvest_doy > 305
+
+```{r}
+
+trait_avg_df <- trait_avg_df %>%
+mutate(
+    planting_stages = case_when(
+      plant_doy <= 120                     ~ "early",
+      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
+      plant_doy > 150                      ~ "late"
+    ) %>% 
+      fct_relevel("early","normal","late"),
+
+    harvest_stages = case_when(
+      harvest_doy <= 275                    ~ "early",
+      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
+      harvest_doy > 305                     ~ "late"
+    ) %>% 
+      fct_relevel("early","normal","late")
+  ) %>%
+  select( -plant_doy, -harvest_doy) %>%
+  ungroup()
+
+trait_avg_df
+
+View(trait_avg_df)
+
+```

### 2025-04-25 code/Draft_script.qmd
d

@@
 
-trait_avg_df
+trait_clean
 
@@
 
-library(stringr)
-
 soil_clean <- soil_df %>%
@@
 
-summary(soil_clean)
+soil_clean
 ```
@@
 
+# Data wrangling on "meta_df"
+
 The following code chunk will conduct data wrangling on  "training_meta.csv".
@@
 
-meta_clean <- meta_df %>%
+meta_df2 <- meta_df %>%
   rename(
@@
 
+meta_df2
+
 ```
@@
 ```{r}
-library(stringr)
 
-meta_clean <- meta_clean %>%
+meta_clean <- meta_df2 %>%
   mutate(
@@
 
+# EDA for cleaned data
+
+```{r}
+summary(trait_clean)
+```
+```{r}
+summary(soil_clean)
+```
+
 ```{r}
@@
 
+# Merging all 3 cleaned data frames
+
 ```{r}
 
-merged_clean  <- trait_avg_df %>%
+merged_clean  <- trait_clean %>%
   left_join(soil_clean, by = c("year", "site")) %>%
@@
 
+merged_clean
+```
+
+```{r}
 summary(merged_clean)
@@
 
-# Exporting
-
-The following code chunk will export the wrangled data into a .csv.
-
-```{r exporting wrangled data set}
+```{r exporting merged data set}
 
@@
 
+The following code chunk will create a map of the USA and plot the sites in the map based on their latitude and longitude.
+
 ```{r create map of USA and add points, message=F, warning=F}
@@
 states <- us_states() %>% 
-  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii) from the rows of "states" object i.e., to keep all states that are NOT "PR", "AK", and "HI" #we use "%in%" to filter (i.e., work on) more than one state/entry #if we wanted to filter just 1 state, we would use == sign e.g., !(state_abbr == c("PR"))
+  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii)
 
@@
 ```
+
+The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.
 
@@
 ```
+The following code chunk will extract unique combinations of year, site, and their coordinates.
 
@@
 
-# Extract unique combinations of year, site, and their coordinates
 site_year_df <- merged_daymet %>%
-  select(year, site, lon, lat) %>%  # include longitude and latitude
-  distinct() %>%                     # one row per unique combination
+  select(year, site, lon, lat) %>%  # need to include longitude and latitude along with site-years
+  distinct() %>%                    
   arrange(year, site)
 
-# Inspect the result
 site_year_df
@@
               values_from = value) %>% 
-  janitor::clean_names() #
+  janitor::clean_names()
 

### 2025-04-25 code/Draft_script.qmd
d

@@
 
+weather_daymet_all
+
 ```
@@
 
+View(weather_daymet_unnest)
+
 ```
@@
 # Merging the weather data retrieved from Daymet to the "merged_clean" data
+
+The following code chunk will merge the weather data retrieved from Daymet to the "merged_clean" data
 
@@
 
-daymet_all_unnest
+fieldweather <- daymet_all_unnest
+
+fieldweather
 
@@
           #"../data/merged_fieldweatherdata.csv"
-          )
+          #)
 
@@
 ```{r}
-#| message: false
-#| warning: false
 
-install.packages("ggridges")
+#install.packages("ggridges")
 
@@
 
+The following code chunk will keep desired variables that we will use further and get abbreviated month name based on the date.
+
 ```{r}
@@
   # Selecting needed variables
-  dplyr::select(year, site, hybrid, lon, lat, yday, 
+  dplyr::select(year, 
+                site, 
+                hybrid, 
+                lon, 
+                lat, 
+                yday, 
                 yield = adjusted_yield, 
-                planting.date = planting_date, 
-                harvest.data = harvest_date, 
-                planting.doy = plant_doy, 
-                harvest.doy = harvest_doy, 
-                season.length = season_len, 
+                prev_crop,
+                plant_doy,
+                harvest_doy,
                 soil.ph = soilpH, 
@@
 
-Now, let's summarize daily weather variables based on month.
+The following code chunk will feature engineer planting and harvest stages based on the following cut-offs:
+
+Planting
+•	early: DOY ≤ 120
+•	normal: 121–150
+•	late: > 150
+
+Harvest
+•	early: DOY ≤ 275
+•	normal: 276–305
+•	late: > 305
+
+```{r}
+
+fe_month <- fe_month%>%
+  mutate(
+    planting_stages = case_when(
+      plant_doy <= 120                     ~ "early",
+      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
+      plant_doy > 150                      ~ "late"
+    ) %>% 
+      fct_relevel("early","normal","late"),
+
+    harvest_stages = case_when(
+      harvest_doy <= 275                    ~ "early",
+      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
+      harvest_doy > 305                     ~ "late"
+    ) %>% 
+      fct_relevel("early","normal","late")
+  ) %>%
+  select(-plant_doy, -plant_doy)
+
+fe_month
+
+```
+
+The following code chunk will summarize daily weather variables based on month.
 
@@
 fe_month_sum <- fe_month %>%
-  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame
+  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame
   #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
@@
 
-Let's check tmax.c and prcp.mm for the first site-year and month.
-
-[Note: the following code chunk is for double checking to make sure that we did everything okay the way we intended]
+The following code chunk will check "tmax.c" and "prcp.mm" for the first site-year and month for the purpose of double checking to make sure that we did everything okay the way we intended.
 
@@
   # Rounding to one decimal point
-  mutate(across(5:ncol(.), ~round(., 1) )) %>%
+  mutate(across(10:ncol(.), ~round(., 1) )) %>%
   select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
@@
 fe_month_sum_wide
+
+View(fe_month_sum_wide)
 

### 2025-04-25 code/Draft_script.qmd
d

@@
 summary(trait_df)
+
+View(trait_df)
 
@@
 
+View(fieldweather)
+
 ```
@@
   mutate(across(10:ncol(.), ~round(., 1) )) %>%
-  select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
+  select(-lon, -lat, -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
 
@@
 ```{r}
-#| message: false
 
@@
 ```
-
-#Export weather_monthsum.csv
 
@@
 
-# Feature engineering "planting stages" and "harvest stages":
+# End of feature engineering
 
-The code chunk below is to feature engineer planting and harvest stages based on the following cut-offs for planting day of the year (plant_doy) and harvest day of the year (harvest_doy):
+# Start of XGboost
 
-Planting stages:
- - early: plant_doy ≤ 120
- - normal: plant_doy = 121–150
- - late: plant_doy > 150
+# ML Workflow with XGBoost
 
-Harvest stages:
- - early: harvest_doy ≤ 275
- - normal: harvest_doy = 276–305
- - late: harvest_doy > 305
+## Step 1: Load Libraries
+
+We begin by loading the required libraries:
+
+```{r}
+#| message: false
+#| warning: false
+#install.packages("xgboost") #new pacakage
+library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
+library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
+library(vip)          # For plotting variable importance from fitted models
+library(xgboost)      # XGBoost implementation in R
+library(ranger)       # Fast implementation of Random Forests
+library(tidyverse)    # Data wrangling and visualization
+library(doParallel)   # For parallel computing (useful during resampling/tuning)
+#library(caret)       # Other great library for Machine Learning 
+```
+
+The following code chunk will rename the "fe_month_sum_wide" data frame as "weather".
 
@@
 
-trait_avg_df <- trait_avg_df %>%
-mutate(
-    planting_stages = case_when(
-      plant_doy <= 120                     ~ "early",
-      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
-      plant_doy > 150                      ~ "late"
-    ) %>% 
-      fct_relevel("early","normal","late"),
-
-    harvest_stages = case_when(
-      harvest_doy <= 275                    ~ "early",
-      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
-      harvest_doy > 305                     ~ "late"
-    ) %>% 
-      fct_relevel("early","normal","late")
-  ) %>%
-  select( -plant_doy, -harvest_doy) %>%
-  ungroup()
-
-trait_avg_df
-
-View(trait_avg_df)
+weather <- fe_month_sum_wide
+weather
 
 ```
+
+# ML workflow
+
+## 1. Pre-processing
+
+```{r weather_split}
+
+set.seed(931735) # Setting seed to get reproducible results
+
+weather_split <- initial_split(
+  weather, 
+  prop = .7, # proption of split same as previous codes
+  strata = yield  # Stratify by target variable
+  )
+
+weather_split
+
+```
+
+### a. Data split
+
+For data split, let's use **70% training / 30% testing**.
+
+```{r weather_train}
+
+weather_train <- training(weather_split)  # 70% of data
+
+weather_train #This is your traing data frame
+
+```
+
+```{r weather_test}
+
+weather_test <- testing(weather_split)    # 30% of data
+
+weather_test
+
+```
+
+### b. Distribution of target variable
+
+```{r distribution}
+ggplot() +
+  geom_density(data = weather_train, 
+               aes(x = yield),
+               color = "red") +
+  geom_density(data = weather_test, 
+               aes(x = yield),
+               color = "blue")
+
+```
+
+### c. Data processing with recipe
+
+Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.
+
+You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html
+
+Different model types require different processing steps.\
+Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data
+
+```{r weather_recipe}
+
+# Create recipe for data preprocessing
+weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
+  step_rm(
+    year,       # Remove year identifier
+    site,       # Remove site identifier
+    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
+  ) %>%
+  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
+  step_dummy(all_nominal_predictors()) %>%
+  # Imputation for any numeric NAs   
+  step_impute_median(all_numeric_predictors())
+
+weather_recipe
+
+```
+
+```{r weather_prep}
+# Prep the recipe to estimate any required statistics
+weather_prep <- weather_recipe %>% 
+  prep()
+
+# Examine preprocessing steps
+weather_prep
+```
+## 2. Training
+
+### a. Model specification
+
+First, let's specify:\
+- the **type of model** we want to train\
+- which **engine** we want to use\
+- which **mode** we want to use
+
+> Elastic nets can only be run for a numerical response variable.
+XgBoost can be run with either numerical (regression) or categorical (classification) explanatory variable.
+Therefore, we have the need to specify the mode here.
+
+XgBoost **hyperparameters**:
+
+-   **`trees`**: The number of boosting rounds (i.e., how many trees will be added sequentially).\
+-   **`tree_depth`**: Controls how deep each individual tree can grow. Deeper trees can capture more complex interactions but also increase the risk of overfitting.\
+-   **`min_n`**: Minimum number of observations required in a node for it to be split. Acts as a regularization tool to prevent overly specific splits.\
+-   **`learn_rate`**: Also known as `eta`, it controls how much each additional tree contributes to the overall model. Smaller values make the model more stable but require more trees.
+
+```{r xgb_spec}
+xgb_spec <- boost_tree(
+  trees = tune(),
+  tree_depth = tune(),  # Maximum depth of each tree
+  min_n = tune(),  # Minimum samples required to split a node
+  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
+  set_engine("xgboost") %>% #specify engine 
+  set_mode("regression")  # Set to mode
+   # Total number of boosting iterations
+
+        # Step size shrinkage for each boosting step
+
+xgb_spec
+
+```
+
+### b. Cross-validation setup
+
+We use 5-fold cross-validation to evaluate model performance during tuning:
+
+```{r}
+
+set.seed(235) #34549
+
+resampling_foldcv <- vfold_cv(weather_train, 
+                              v = 5)
+
+resampling_foldcv
+
+resampling_foldcv$splits[[1]]
+
+```
+### c. Hyperparameter grid with Latin Hypercube Sampling
+
+We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:
+
+```{r }
+
+xgb_grid <- grid_latin_hypercube(
+  tree_depth(),
+  min_n(),
+  learn_rate(),
+  trees(),
+  size = 10
+)
+
+xgb_grid
+```
+
+```{r}
+ggplot(data = xgb_grid,
+       aes(x = tree_depth, 
+           y = min_n)) +
+  geom_point(aes(color = factor(learn_rate), #coloring the bubbles based on learn_rate
+                 size = trees), #size of the bubbles are based on the tress
+             alpha = .5,
+             show.legend = FALSE)
+```
+
+## 3. Model Tuning
+
+```{r xgb_grid_result}
+
+#install.packages("doParallel")
+#install.packages("parallel")
+
+library(doParallel)
+library(parallel)
+
+set.seed(76544)
+
+#parallel processing
+registerDoParallel(cores = parallel::detectCores()-1)
+
+xgb_res <- tune_race_anova(object = xgb_spec,
+                      preprocessor = weather_recipe,
+                      resamples = resampling_foldcv,
+                      grid = xgb_grid,
+                      control = control_race(save_pred = TRUE))
+
+stopImplicitCluster()
+
+beepr::beep()
+
+xgb_res$.metrics[[2]]
+
+```

### 2025-04-26 code/Draft_script.qmd
d

@@
 #install.packages("xgboost") #new pacakage
+#install.packages("caret")
+
 library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
@@
 library(doParallel)   # For parallel computing (useful during resampling/tuning)
-#library(caret)       # Other great library for Machine Learning 
+library(caret)       # Other great library for Machine Learning 
 ```
@@
 # Create recipe for data preprocessing
-weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
+weather_recipe <- recipe(yield ~ ., data = weather_train) %>% 
+  # Remove identifier columns and months not in growing season
   step_rm(
@@
     site,       # Remove site identifier
+    hybrid,     # Remove site identifier
     matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
@@
   # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
-  step_dummy(all_nominal_predictors()) %>%
-  # Imputation for any numeric NAs   
-  step_impute_median(all_numeric_predictors())
+  step_dummy(all_nominal_predictors())
 
@@
 ```{r xgb_spec}
+
 xgb_spec <- boost_tree(
@@
   set_mode("regression")  # Set to mode
-   # Total number of boosting iterations
-
-        # Step size shrinkage for each boosting step
 
@@
 ```
+
+## 4. Select Best Models
+
+We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:
+
+```{r}
+# Based on lowest RMSE
+best_rmse <- xgb_res %>% 
+  select_best(metric = "rmse")%>% 
+  mutate(source = "best_rmse")
+
+best_rmse
+
+```
+
+```{r}
+# Based on lowers RMSE within 1% loss
+best_rmse_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rmse",
+                     limit = 1
+                     )%>% 
+  mutate(source = "best_rmse_pct_loss")
+
+best_rmse_pct_loss
+```
+
+```{r}
+# Based on lowest RMSE within 1 se
+best_rmse_one_std_err <- xgb_res %>% 
+  select_by_one_std_err(metric = "rmse",
+                        eval_time = 100,
+                        trees
+                        )%>% 
+  mutate(source = "best_rmse_one_std_err")
+
+best_rmse_one_std_err
+```
+
+Here we use all three methods which we learn in this class for R2.
+
+```{r}
+# Based on greatest R2
+best_r2 <- xgb_res %>% 
+  select_best(metric = "rsq")%>% 
+  mutate(source = "best_r2")
+
+best_r2
+```
+
+```{r}
+# Based on lowers R2 within 1% loss
+best_r2_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rsq",
+                     limit = 1
+                     ) %>% 
+  mutate(source = "best_r2_pct_loss")
+
+best_r2_pct_loss
+```
+
+```{r}
+# Based on lowest R2 within 1 se
+best_r2_one_std_error <- xgb_res %>% 
+  select_by_one_std_err(metric = "rsq",
+                        eval_time = 100,
+                        trees
+                        ) %>%
+  mutate(source = "best_r2_one_std_error")
+
+best_r2_one_std_error
+```
+
+## Compare and Finalize Model
+
+```{r comparing values}
+best_rmse %>% 
+  bind_rows(best_rmse_pct_loss, 
+            best_rmse_one_std_err, 
+            best_r2, 
+            best_r2_pct_loss, 
+            best_r2_one_std_error)
+```
+
+## 5. Final Specification
+
+```{r final_spec_fit}
+final_spec <- boost_tree(
+  trees = best_r2$trees,           # Number of boosting rounds (trees)
+  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
+  min_n = best_r2$min_n,           # Minimum number of samples to split a node
+  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+final_spec
+```
+
+## 6. Final Fit and Predictions
+
+## Validation
+
+```{r final_fit}
+set.seed(10)
+final_fit <- last_fit(final_spec,
+                weather_recipe,
+                split = weather_split)
+
+final_fit %>%
+  collect_predictions()
+```
+
+## 7. Evaluate on *Test Set*
+
+```{r final_fit_metrics}
+final_fit %>%
+  collect_metrics()
+```
+
+-R2 & RMSE =
+
+-best_r2_pct_loss =
+
+-best_r2_one_std_error =
+
+-best_rmse_pct_loss =
+
+-best_rmse_one_std_err =
+
+## 8. Evaluate on Training Set
+
+```{r}
+final_spec %>%
+  fit(strength_gtex ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rmse(strength_gtex, .pred) %>%
+  bind_rows(
+
+# R2
+final_spec %>%
+  fit(strength_gtex ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rsq(strength_gtex, .pred))
+```
+
+## 9. Predicted vs Observed Plot
+
+```{r}
+final_fit %>%
+  collect_predictions() %>%
+  ggplot(aes(x = strength_gtex,
+             y = .pred)) +
+  geom_point() +
+  geom_abline() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous(limits = c(20, 40)) +
+  scale_y_continuous(limits = c(20, 40)) 
+```
+
+## 10. Variable Importance
+
+```{r final_spec}
+final_spec %>%
+  fit(strength_gtex ~ .,
+         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
+    vi() %>%
+  mutate(
+    Variable = fct_reorder(Variable, 
+                           Importance)
+  ) %>%
+  ggplot(aes(x = Importance, 
+             y = Variable)) +
+  geom_col() +
+  scale_x_continuous(expand = c(0, 0)) +
+  labs(y = NULL)
+```
+
+# End of XGboost
+
+# Start of Random Forest
+
+# Setup
+
+```{r}
+#| message: false
+#| warning: false
+
+#install.packages("ranger") #one of the packages in R that models random forest
+
+library(tidymodels)
+library(tidyverse) #takes a while to load
+library(vip)
+library(ranger)
+library(finetune)
+```
+
+## 2. Random Forest Training  
+### a. Random Forest Model specification
+
+First, let's specify:  
+  - the **type of model** we want to train  
+  - which **engine** we want to use  
+  - which **mode** we want to use
+
+Random forest **hyperparameters**:  
+  - **trees**: number of trees in the forest    
+  - **mtry**: number of random features [or columns] sampled at each node split    
+    - **min_n**: minimum number of data points in a node that are required for the node to be split further
+
+[Important note: we are not gonna fine tune "min_n" in this exercise because of time constraint, but we should fine tune "min_n" to do random forest in our own data]
+
+Let's create a model specification that will **fine-tune** the first two for us.
+
+A given model type can be fit with different engines (e.g., through different packages). Here, we'll use the **ranger** engine/package.
+
+```{r rf_spec}
+
+rf_spec <- 
+  # Specifying random forest as our model type, asking to tune the hyperparameters
+  rand_forest(trees = tune(),
+              mtry = tune() #both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
+              ) %>%
+    # Specify the engine (= package)
+    set_engine("ranger") %>% #specifying "ranger" as the engine/package to run random forest 
+    # Specifying mode  
+    set_mode("regression") #random forest can handle both regression (when y is numerical) and classification (when y is categorical) #Here, we are specifying "set_mode("regression")" because our y variable is numerical [continuous]
+
+rf_spec
+
+```
+
+### b. Hyper-parameter tuning
+
+For our iterative search, we need:  
+  - Our model specification (`rf_spec`)  
+  - The recipe (`weather_recipe`)  
+  - Our **resampling strategy** (don't have yet)
+
+Let's define our resampling strategy below, using a 5-fold cross validation approach:
+
+```{r resampling_foldcv}
+
+set.seed(34549)
+
+rf_resampling_foldcv <- vfold_cv(weather_train, 
+                              v = 5) #but at least 10 folds are recommended
+
+rf_resampling_foldcv
+rf_resampling_foldcv$splits[[1]]
+rf_resampling_foldcv$splits[[2]]
+
+```
+
+We will use an iterative search algorithm called **simulated annealing**.
+
+In the algorithm below, we are asking for 10 iterations.
+
+```{r rf_grid_result}
+
+set.seed(76544)
+
+rf_grid_result <- tune_sim_anneal(object = rf_spec,
+                     preprocessor = weather_recipe,
+                     resamples = resampling_foldcv,
+                    #param_info = rf_param,
+                    iter = 10 
+                     )
+
+beepr::beep()
+
+rf_grid_result
+rf_grid_result$.metrics[[2]]
+```
+
+Let's collect a summary of metrics (across all folds, for each iteration), and plot them.
+
+First, RMSE (lower is better):
+
+```{r RMSE}
+rf_grid_result %>%
+  collect_metrics() %>%
+  filter(.metric == "rmse") %>%
+  ggplot(aes(x = mtry, 
+             y = trees 
+             )) +
+  geom_path(group = 1) +
+  geom_point(aes(color = mean),
+             size = 3) + 
+  scale_color_viridis_b() +
+  geom_text(aes(label = .iter), nudge_x = .0005) +
+  labs(title = "RMSE")
+```
+
+We want to minimize the RMSE.
+The lower the RMSE, the better.
+In the legend section of the RMSE grid plot seen above, the lowest RMSE is"2.85" which has a dark purple color, and we see the dark purple colored dot is on 4 th iteration, which indicates that the lowest RMSE [= "2.85"] was obtained in the 4 th iteration.
+
+What tree_depth and mtry values created lowest RMSE?
+
+tree_depth = ~ 110; mtry = ~ 22 [when we have 10 iterations]
+
+Now, let's look into R2 (higher is better):
+
+```{r R2}
+rf_grid_result %>%
+  collect_metrics() %>%
+  filter(.metric == "rsq") %>%
+  ggplot(aes(x = mtry, 
+             y = trees 
+             )) +
+  geom_path(group = 1) +
+  geom_point(aes(color = mean),
+             size = 3) + 
+  scale_color_viridis_b() +
+  geom_text(aes(label = .iter), nudge_x = .0005) +
+  labs(title = "R2")
+
+```
+
+We want to see higher R2 (R2: higher is better).
+
+We want to see a maximized R2.
+The higher the R2, the better.
+In the legend section of the R2 grid plot seen above, the highest R2 is "0.16" which has a yellow color, and we see the yellow colored dot is on 4 th iteration, which indicates that the highest R2 [= "0.16"] was obtained in the 4 th iteration.
+
+What tree_depth and min criterion values created highest R2?
+
+tree_depth = ~110; mtry = ~ 22 [this was the best result for 10 iterations]
+
+> Previously, we selected the single best model. Now, let's select the best model by "percent loss" of the metric, so we choose a model among the top ones that is more parsimonious.
+
+```{r}
+# Based on lowest RMSE
+best_rmse <- rf_grid_result %>%
+  select_by_pct_loss("trees",
+                     metric = "rmse",
+                     limit = 2 #"limit = 2": this is how many percent losses you are accepting, 2% in this case
+                     )
+
+best_rmse
+
+```
+
+```{r}
+# Based on greatest R2
+best_r2 <- rf_grid_result %>%
+  select_by_pct_loss("trees",
+                     metric = "rsq",
+                     limit = 2
+                     )
+
+best_r2
+
+```
+
+I got the he following results that were different than Dr. Bastos's results (because of the randomness of 2 hyperparameters going on at each tree for a lot of models):
+
+Based on RMSE, we would choose   
+  - mtry = 25 
+  - trees = 443
+
+Based on R2, we would choose   
+  - mtry = 22
+  - trees = 113
+
+Let's use the hyperparameter values that optimized R2 to fit our final model.
+
+```{r final_spec}
+final_spec <- rand_forest(trees = best_r2$trees,
+                          mtry = best_r2$mtry) %>%
+  # Specify the engine
+  set_engine("ranger",
+             importance = "permutation" #"permutation" is how we look into variable importance in random forest
+             ) %>%
+    # Specifying mode  
+  set_mode("regression")
+
+final_spec
+```
+
+## 3. Validation
+
+Now that we determined our best model, let's do our **last fit**.
+
+### What does last fit mean? (Will come in exam)
+
+This means 2 things:  
+  - Training the optimum hyperparameter values on the **entire training set**  
+  - Using it to **predict** on the **test set**
+
+These 2 steps can be completed in one function, as below:
+
+```{r final_fit}
+final_fit <- last_fit(final_spec, 
+                weather_recipe,
+                split = weather_split)
+
+final_fit %>%
+  collect_predictions()
+```
+
+Metrics on the **test set**:
+
+```{r}
+final_fit %>%
+  collect_metrics()
+```
+
+From the output,
+Fit metrics on the test set:
+rmse = 2.9317445	
+rsq	= 0.1882245
+
+Metrics on **train set** (for curiosity and compare to test set):  
+```{r}
+# RMSE
+final_spec %>%
+  fit(strength_gtex ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rmse(strength_gtex, .pred) %>%
+  bind_rows(
+
+    # R2
+    final_spec %>%
+      fit(strength_gtex ~ .,
+          data = bake(weather_prep, 
+                      weather_train)) %>%
+      augment(new_data = bake(weather_prep, 
+                              weather_train)) %>% 
+      rsq(strength_gtex, .pred)
+
+  )
+
+```
+
+From the output,
+Fit metrics on the train set:
+rmse = 1.1385078	
+rsq	= 0.9470296
+
+How does metrics on test compare to metrics on train?
+
+Fit metrics on the test set:
+rmse = 2.9317445	
+rsq	= 0.1882245
+
+Fit metrics on the train set:
+rmse = 1.1385078	
+rsq	= 0.9470296
+
+We see that the R2 is really r-e-a-l-l-y high in the train set (rsq	= 0.9470296), as compared to the R2 of the test set 0.1882245.
+The RMSE is lower in the train set (rmse = 1.1385078) as compared to the RMSE of the test set (rmse = 2.9317445).
+[This is the reason why it's okay to report the fit metrics of the train set to understand how well our model is explaining our data, but it is not a good measure of the predictive ability of the model.
+To understand the predictive ability of the model, we have to look at the fit metrics of the test set.
+Notice that the R2 of the test set was 0.1882245, whereas the R2 of the train set was 0.9470296 -- which is a very large difference.
+Hence, *we need to always report the fit metrics of the test set* along with the fit metrics of the train set.
+]
+
+Predicted vs. observed plot:  
+```{r}
+final_fit %>%
+  collect_predictions() %>%
+  ggplot(aes(x = strength_gtex,
+             y = .pred)) +
+  geom_point() +
+  geom_abline() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous(limits = c(20, 40)) +
+  scale_y_continuous(limits = c(20, 40)) 
+```
+
+[**Very important note:**
+To interpret random forest model, we should always expcet to see (1) *perdicted vs.
+observed plot* and (3) *variable importance plot*, at the least -- these are the 2 things that must be mentioned in a random forest model.
+]
+
+*Variable importance:*
+
+[For Random Forest,] The importance metric we are evaluating here is **permutation**.
+
+In the permutation-based approach, for each tree, the out-of-bag sample is passed down the tree and the prediction accuracy is recorded. [as is]
+
+Then the values for each variable (one at a time) are randomly permuted and the accuracy is again computed.
+
+The decrease in accuracy as a result of this randomly shuffling of feature values is averaged over all the trees for each predictor.
+
+The variables with the **largest average decrease in accuracy** [after being permuted] are [the ones that are] considered **most important**.
+
+```{r}
+
+final_spec %>%
+  fit(strength_gtex ~ .,
+         data = bake(weather_prep, weather)) %>%
+    vi() %>%
+  mutate(
+    Variable = fct_reorder(Variable, 
+                           Importance)
+  ) %>%
+  ggplot(aes(x = Importance, 
+             y = Variable)) +
+  geom_col() +
+  scale_x_continuous(expand = c(0, 0)) +
+  labs(y = NULL)
+
+```
+
+#End of Random Forest
+
+#Start of getting yield prediction (yield_mg_ha) for the unseen data in "testing_submission.csv" file

### 2025-04-26 code/Draft_script.qmd
d

@@
 ## 2. Random Forest Training
+
 ### a. Random Forest Model specification
@@
                      preprocessor = weather_recipe,
-                     resamples = resampling_foldcv,
+                     resamples = rf_resampling_foldcv,
                     #param_info = rf_param,
@@
 
-We want to minimize the RMSE.
-The lower the RMSE, the better.
-In the legend section of the RMSE grid plot seen above, the lowest RMSE is"2.85" which has a dark purple color, and we see the dark purple colored dot is on 4 th iteration, which indicates that the lowest RMSE [= "2.85"] was obtained in the 4 th iteration.
-
-What tree_depth and mtry values created lowest RMSE?
-
-tree_depth = ~ 110; mtry = ~ 22 [when we have 10 iterations]
-
 Now, let's look into R2 (higher is better):
@@
 ```{r R2}
+
 rf_grid_result %>%
@@
 
-We want to see a maximized R2.
-The higher the R2, the better.
-In the legend section of the R2 grid plot seen above, the highest R2 is "0.16" which has a yellow color, and we see the yellow colored dot is on 4 th iteration, which indicates that the highest R2 [= "0.16"] was obtained in the 4 th iteration.
-
-What tree_depth and min criterion values created highest R2?
-
-tree_depth = ~110; mtry = ~ 22 [this was the best result for 10 iterations]
-
-> Previously, we selected the single best model. Now, let's select the best model by "percent loss" of the metric, so we choose a model among the top ones that is more parsimonious.
-
-```{r}
+The following code chunk will select the best model based on "percent loss" of the metric.
+
+```{r}
+
 # Based on lowest RMSE
@@
 
-```{r}
+The following code chunk will select the best model based on greatest "R2"
+
+```{r}
+
 # Based on greatest R2

### 2025-04-26 code/Draft_script.qmd
d

@@
 fe_month_sum <- fe_month %>%
-  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame
+  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame
   #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
@@
 final_spec %>%
-  fit(strength_gtex ~ .,
+  fit(yield ~ .,
       data = bake(weather_prep, 
@@
                           weather_train)) %>% 
-  rmse(strength_gtex, .pred) %>%
+  rmse(yield, .pred) %>%
   bind_rows(
@@
                           weather_train)) %>% 
-  rsq(strength_gtex, .pred))
+  rsq(yield, .pred))
 ```
@@
   collect_predictions() %>%
-  ggplot(aes(x = strength_gtex,
+  ggplot(aes(x = yield,
              y = .pred)) +
@@
 final_spec %>%
-  fit(strength_gtex ~ .,
+  fit(yield ~ .,
          data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
@@
     final_spec %>%
-      fit(strength_gtex ~ .,
+      fit(yield ~ .,
           data = bake(weather_prep, 
@@
                               weather_train)) %>% 
-      rsq(strength_gtex, .pred)
+      rsq(yield, .pred)
 
@@
 final_spec %>%
-  fit(strength_gtex ~ .,
+  fit(yield ~ .,
          data = bake(weather_prep, weather)) %>%

### 2025-04-26 code/Draft_script.qmd
d

@@
 The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.
+
+Declaration of AI use: the following code chunk was inspired and subsequently modified on the basis of code initially generated by ChatGPT
 
@@
   trees = tune(),
-  tree_depth = tune(),  # Maximum depth of each tree
-  min_n = tune(),  # Minimum samples required to split a node
-  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
+  tree_depth = tune() # ,  # Maximum depth of each tree
+  #min_n = tune(),  # Minimum samples required to split a node
+  #learn_rate = tune()
+  ) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
   set_engine("xgboost") %>% #specify engine 
@@
 xgb_grid <- grid_latin_hypercube(
+  trees(),
   tree_depth(),
-  min_n(),
-  learn_rate(),
-  trees(),
+  #min_n(),
+  #learn_rate(),
   size = 10

### 2025-04-26 code/Draft_script.qmd
d

@@
 
+Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.
+
 ```{r}
@@
 #parallel processing
-registerDoParallel(cores = parallel::detectCores()-1)
+registerDoParallel(cores = parallel::detectCores()-1) #starts parallel processing
 
@@
 
-stopImplicitCluster()
+stopImplicitCluster() #ends parallel processing
 
@@
   trees = best_r2$trees,           # Number of boosting rounds (trees)
-  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
-  min_n = best_r2$min_n,           # Minimum number of samples to split a node
-  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
+  tree_depth = best_r2$tree_depth #, # Maximum depth of each tree
+  #min_n = best_r2$min_n,           # Minimum number of samples to split a node
+  #learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
 ) %>%
@@
 ```
-
--R2 & RMSE =
-
--best_r2_pct_loss =
-
--best_r2_one_std_error =
-
--best_rmse_pct_loss =
-
--best_rmse_one_std_err =
 
@@
   geom_smooth(method = "lm") +
-  scale_x_continuous(limits = c(20, 40)) +
-  scale_y_continuous(limits = c(20, 40)) 
+  scale_x_continuous() +
+  scale_y_continuous() 
 ```
@@
   rand_forest(trees = tune(),
-              mtry = tune() #both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
+              mtry = tune(),
+              min_n = tune()#both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
               ) %>%

### 2025-04-26 code/Draft_script.qmd
d

@@
 
-## 3. Model Tuning
+## 3. XGboost Model Tuning
 
@@
 # End of XGboost
-
-# Start of Random Forest
-
-# Setup
-
-```{r}
-#| message: false
-#| warning: false
-
-#install.packages("ranger") #one of the packages in R that models random forest
-
-library(tidymodels)
-library(tidyverse) #takes a while to load
-library(vip)
-library(ranger)
-library(finetune)
-```
-
-## 2. Random Forest Training
-
-### a. Random Forest Model specification
-
-First, let's specify:  
-  - the **type of model** we want to train  
-  - which **engine** we want to use  
-  - which **mode** we want to use
-
-Random forest **hyperparameters**:  
-  - **trees**: number of trees in the forest    
-  - **mtry**: number of random features [or columns] sampled at each node split    
-    - **min_n**: minimum number of data points in a node that are required for the node to be split further
-
-[Important note: we are not gonna fine tune "min_n" in this exercise because of time constraint, but we should fine tune "min_n" to do random forest in our own data]
-
-Let's create a model specification that will **fine-tune** the first two for us.
-
-A given model type can be fit with different engines (e.g., through different packages). Here, we'll use the **ranger** engine/package.
-
-```{r rf_spec}
-
-rf_spec <- 
-  # Specifying random forest as our model type, asking to tune the hyperparameters
-  rand_forest(trees = tune(),
-              mtry = tune(),
-              min_n = tune()#both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
-              ) %>%
-    # Specify the engine (= package)
-    set_engine("ranger") %>% #specifying "ranger" as the engine/package to run random forest 
-    # Specifying mode  
-    set_mode("regression") #random forest can handle both regression (when y is numerical) and classification (when y is categorical) #Here, we are specifying "set_mode("regression")" because our y variable is numerical [continuous]
-
-rf_spec
-
-```
-
-### b. Hyper-parameter tuning
-
-For our iterative search, we need:  
-  - Our model specification (`rf_spec`)  
-  - The recipe (`weather_recipe`)  
-  - Our **resampling strategy** (don't have yet)
-
-Let's define our resampling strategy below, using a 5-fold cross validation approach:
-
-```{r resampling_foldcv}
-
-set.seed(34549)
-
-rf_resampling_foldcv <- vfold_cv(weather_train, 
-                              v = 5) #but at least 10 folds are recommended
-
-rf_resampling_foldcv
-rf_resampling_foldcv$splits[[1]]
-rf_resampling_foldcv$splits[[2]]
-
-```
-
-We will use an iterative search algorithm called **simulated annealing**.
-
-In the algorithm below, we are asking for 10 iterations.
-
-```{r rf_grid_result}
-
-set.seed(76544)
-
-rf_grid_result <- tune_sim_anneal(object = rf_spec,
-                     preprocessor = weather_recipe,
-                     resamples = rf_resampling_foldcv,
-                    #param_info = rf_param,
-                    iter = 10 
-                     )
-
-beepr::beep()
-
-rf_grid_result
-rf_grid_result$.metrics[[2]]
-```
-
-Let's collect a summary of metrics (across all folds, for each iteration), and plot them.
-
-First, RMSE (lower is better):
-
-```{r RMSE}
-rf_grid_result %>%
-  collect_metrics() %>%
-  filter(.metric == "rmse") %>%
-  ggplot(aes(x = mtry, 
-             y = trees 
-             )) +
-  geom_path(group = 1) +
-  geom_point(aes(color = mean),
-             size = 3) + 
-  scale_color_viridis_b() +
-  geom_text(aes(label = .iter), nudge_x = .0005) +
-  labs(title = "RMSE")
-```
-
-Now, let's look into R2 (higher is better):
-
-```{r R2}
-
-rf_grid_result %>%
-  collect_metrics() %>%
-  filter(.metric == "rsq") %>%
-  ggplot(aes(x = mtry, 
-             y = trees 
-             )) +
-  geom_path(group = 1) +
-  geom_point(aes(color = mean),
-             size = 3) + 
-  scale_color_viridis_b() +
-  geom_text(aes(label = .iter), nudge_x = .0005) +
-  labs(title = "R2")
-
-```
-
-We want to see higher R2 (R2: higher is better).
-
-The following code chunk will select the best model based on "percent loss" of the metric.
-
-```{r}
-
-# Based on lowest RMSE
-best_rmse <- rf_grid_result %>%
-  select_by_pct_loss("trees",
-                     metric = "rmse",
-                     limit = 2 #"limit = 2": this is how many percent losses you are accepting, 2% in this case
-                     )
-
-best_rmse
-
-```
-
-The following code chunk will select the best model based on greatest "R2"
-
-```{r}
-
-# Based on greatest R2
-best_r2 <- rf_grid_result %>%
-  select_by_pct_loss("trees",
-                     metric = "rsq",
-                     limit = 2
-                     )
-
-best_r2
-
-```
-
-I got the he following results that were different than Dr. Bastos's results (because of the randomness of 2 hyperparameters going on at each tree for a lot of models):
-
-Based on RMSE, we would choose   
-  - mtry = 25 
-  - trees = 443
-
-Based on R2, we would choose   
-  - mtry = 22
-  - trees = 113
-
-Let's use the hyperparameter values that optimized R2 to fit our final model.
-
-```{r final_spec}
-final_spec <- rand_forest(trees = best_r2$trees,
-                          mtry = best_r2$mtry) %>%
-  # Specify the engine
-  set_engine("ranger",
-             importance = "permutation" #"permutation" is how we look into variable importance in random forest
-             ) %>%
-    # Specifying mode  
-  set_mode("regression")
-
-final_spec
-```
-
-## 3. Validation
-
-Now that we determined our best model, let's do our **last fit**.
-
-### What does last fit mean? (Will come in exam)
-
-This means 2 things:  
-  - Training the optimum hyperparameter values on the **entire training set**  
-  - Using it to **predict** on the **test set**
-
-These 2 steps can be completed in one function, as below:
-
-```{r final_fit}
-final_fit <- last_fit(final_spec, 
-                weather_recipe,
-                split = weather_split)
-
-final_fit %>%
-  collect_predictions()
-```
-
-Metrics on the **test set**:
-
-```{r}
-final_fit %>%
-  collect_metrics()
-```
-
-From the output,
-Fit metrics on the test set:
-rmse = 2.9317445	
-rsq	= 0.1882245
-
-Metrics on **train set** (for curiosity and compare to test set):  
-```{r}
-# RMSE
-final_spec %>%
-  fit(yield ~ .,
-      data = bake(weather_prep, 
-                  weather_train)) %>%
-  augment(new_data = bake(weather_prep, 
-                          weather_train)) %>% 
-  rmse(yield, .pred) %>%
-  bind_rows(
-
-    # R2
-    final_spec %>%
-      fit(yield ~ .,
-          data = bake(weather_prep, 
-                      weather_train)) %>%
-      augment(new_data = bake(weather_prep, 
-                              weather_train)) %>% 
-      rsq(yield, .pred)
-
-  )
-
-```
-
-From the output,
-Fit metrics on the train set:
-rmse = 1.1385078	
-rsq	= 0.9470296
-
-How does metrics on test compare to metrics on train?
-
-Fit metrics on the test set:
-rmse = 2.9317445	
-rsq	= 0.1882245
-
-Fit metrics on the train set:
-rmse = 1.1385078	
-rsq	= 0.9470296
-
-We see that the R2 is really r-e-a-l-l-y high in the train set (rsq	= 0.9470296), as compared to the R2 of the test set 0.1882245.
-The RMSE is lower in the train set (rmse = 1.1385078) as compared to the RMSE of the test set (rmse = 2.9317445).
-[This is the reason why it's okay to report the fit metrics of the train set to understand how well our model is explaining our data, but it is not a good measure of the predictive ability of the model.
-To understand the predictive ability of the model, we have to look at the fit metrics of the test set.
-Notice that the R2 of the test set was 0.1882245, whereas the R2 of the train set was 0.9470296 -- which is a very large difference.
-Hence, *we need to always report the fit metrics of the test set* along with the fit metrics of the train set.
-]
-
-Predicted vs. observed plot:  
-```{r}
-final_fit %>%
-  collect_predictions() %>%
-  ggplot(aes(x = yield,
-             y = .pred)) +
-  geom_point() +
-  geom_abline() +
-  geom_smooth(method = "lm") +
-  scale_x_continuous(limits = c(20, 40)) +
-  scale_y_continuous(limits = c(20, 40)) 
-```
-
-[**Very important note:**
-To interpret random forest model, we should always expcet to see (1) *perdicted vs.
-observed plot* and (3) *variable importance plot*, at the least -- these are the 2 things that must be mentioned in a random forest model.
-]
-
-*Variable importance:*
-
-[For Random Forest,] The importance metric we are evaluating here is **permutation**.
-
-In the permutation-based approach, for each tree, the out-of-bag sample is passed down the tree and the prediction accuracy is recorded. [as is]
-
-Then the values for each variable (one at a time) are randomly permuted and the accuracy is again computed.
-
-The decrease in accuracy as a result of this randomly shuffling of feature values is averaged over all the trees for each predictor.
-
-The variables with the **largest average decrease in accuracy** [after being permuted] are [the ones that are] considered **most important**.
-
-```{r}
-
-final_spec %>%
-  fit(yield ~ .,
-         data = bake(weather_prep, weather)) %>%
-    vi() %>%
-  mutate(
-    Variable = fct_reorder(Variable, 
-                           Importance)
-  ) %>%
-  ggplot(aes(x = Importance, 
-             y = Variable)) +
-  geom_col() +
-  scale_x_continuous(expand = c(0, 0)) +
-  labs(y = NULL)
-
-```
-
-#End of Random Forest
-
-#Start of getting yield prediction (yield_mg_ha) for the unseen data in "testing_submission.csv" file
